Developing effective theories for single and bilayer graphene using density matrix downfolding

### 2019-09-08 draft.tex
added extra refs to intro + rewrote abstract

\begin{document}
\maketitle
\begin{abstract}
A core problem of condensed matter physics is understanding how to develop low-energy effective models for realistic systems.
I believe that a density matrix downfolding (DMD) method \cite{Zheng2017} can be used to systematically build effective theories for such systems beginning with a high energy theory like the \textit{ab-initio} Hamiltonian.
I present as a preliminary result an effective Hamiltonian developed using DMD which can accurately reproduce the spectra and properties of the lowest lying excitations of the CuO molecule.
I propose using DMD to develop accurate model Hamiltonians for single and bilayer graphene.
My work will serve to mature the DMD method and take steps towards a model for the strongly correlated twisted bilayer graphene system.
\end{abstract}
Committee: Lucas Wagner, Brian DeMarco, Andre Schleife, Karin Dahmen
\pagebreak

@@

Yet unresolved is a paradigm for building many-body model Hamiltonians which can describe strongly correlated systems.
Band and Fermi liquid theory have been successful in accurately modeling systems where effective interactions are negligible at low energy scales.
These theories, however, fail in describing the low-energy excitations of strongly correlated systems like high Tc superconductors and twisted bilayer graphene (TBLG).
Many-body models like the Hubbard \cite{Hubbard1963}, Kanamori \cite{10.1143/PTP.30.275}, t-J \cite{Chao_1977} and Heisenberg models have been proposed to describe such strongly correlated systems, but their effectiveness in describing realistic materials is unclear.

In an attempt to build accurate many-body Hamiltonians for realistic systems, a framework for model development using solutions from \textit{ab-initio} simulations has emerged.
Common approaches involve density functional theory (DFT) \cite{PhysRev.140.A1133} with one-body terms obtained from the DFT band structure \cite{PhysRevLett.87.047003} and two-body terms developed using constrained DFT, RPA or other methods \cite{PhysRevB.62.R16219, PhysRevB.70.195104, PhysRevB.88.075106}.
L\"owdin methods and canonical transformations have also been used to develop effective theories \cite{doi:10.1063/1.4802766, PhysRevA.81.013402, doi:10.1063/1.2196410}.
While useful, it is typically impossible to know whether models built from these techniques accurately describe reality, as measures of model quality are rarely reported.
Instead, out-of-method comparisons of model predictions to experiment must be conducted for model validation.

A recently developed density matrix downfolding (DMD) method provides a more direct link between \textit{ab-initio} and many-body effective Hamiltonians.
Similar to the methods above, DMD uses information from \textit{ab-initio} simulations on the system to fit the best parameters for a given many-body model \textit{ansatz}.
Unlike the methods above, the resultant fit model is accompanied by a clear measure of model quality, quantifying the ability of the model to reproduce the low-energy spectrum of the system.
Since model quality is reported, DMD allows for direct comparison between different model \textit{ansatzes} as well, an ability which can be used to learn the most important interaction terms for a given system directly from \textit{ab-initio} calculations.
For example, DMD calculations on benzene and single layer graphene determined that long-range coulomb interactions are necessary in modeling both systems.

The utility of DMD is made apparent when studying the recent surge in model development for twisted bilayer graphene (TBLG).
\textit{Ab-initio} model development has so far been restricted to tight-binding models which accurately describe single particle effects like band flattening, but without a report of model quality cannot determine the role of interactions in the system.
Minimal interacting models have been proposed using traditional theoretical methods, but without a link to \textit{ab-initio} it is unclear whether these simplified models can capture the complexity of the real system.
Starting with the tight-binding \textit{ansatz}, DMD can be used to learn the most important interactions in the system, resulting in a final many-body model which can accurately describe TBLG.

As such, one portion of a project between research groups at UIUC, UCSB and Rice University is dedicated towards the development of an accurate interacting model for TBLG.
The model development consists of two layers of downfolding: first, developing an atomic scale lattice model for TBLG using \textit{ab-initio} simulations, and second, downfolding the resultant lattice model to a continuum length scale.
My thesis will be focused on the first layer of downfolding, which I will conduct within the DMD framework using accurate \textit{ab-initio} fixed-node diffusion Monte Carlo calculations.
I propose a two year plan which will result in atomic scale, many-body lattice Hamiltonians for single- (SLG) and simple bi-layer (BLG) graphene systems.
My work will lay the necessary ground work for the development of a fully twist angle dependent lattice model for TBLG, and the eventual continuum model described above.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Building accurate model Hamiltonians for realistic materials is a fundamental problem in modern condensed matter physics.
A recently developed density matrix downfolding (DMD) \cite{Zheng2017} method provides a very direct framework for downfolding the \textit{ab-initio} Hamiltonian to a many-body effective Hamiltonian using only \textit{ab-initio} calculations. 
Here I propose a two year thesis plan for developing atomic scale many-body model Hamiltonians for single- and simple bi-layer graphene systems within the DMD framework using high accuracy \textit{ab-initio} quantum Monte Carlo calculations.
The results of my thesis will lay the groundwork for a continuing four year project to build a twist dependent many-body continuum model for the twisted bilayer graphene (TBLG) system.
To demonstrate my preparedness I present a model Hamiltonian developed using DMD which accurately reproduces the spectra and properties of the lowest lying excitations of the CuO molecule.
\end{abstract}
Committee: Lucas Wagner, Brian DeMarco, Andre Schleife, Karin Dahmen
\pagebreak

@@

Yet unresolved is a paradigm for building many-body model Hamiltonians which can describe strongly correlated systems.
Band and Fermi liquid theory have been successful in accurately modeling systems where effective interactions are negligible at low energy scales.
These theories, however, fail in describing the low-energy excitations of strongly correlated systems like high Tc superconductors and TBLG.
Minimal many-body models like the Hubbard \cite{Hubbard1963}, Kanamori \cite{10.1143/PTP.30.275}, t-J \cite{Chao_1977} and Heisenberg models have been proposed to describe such strongly correlated systems, their effectiveness in describing realistic materials is unclear.

In an attempt to build accurate many-body Hamiltonians for realistic systems, a framework for model development using solutions from \textit{ab-initio} simulations has emerged.
Common approaches involve density functional theory (DFT) \cite{PhysRev.140.A1133} with one-body terms obtained from the DFT band structure \cite{PhysRevLett.87.047003} and two-body terms developed using constrained DFT, random phase approximation (RPA) or other methods \cite{PhysRevB.62.R16219, PhysRevB.70.195104, PhysRevB.88.075106}.
L\"owdin methods and canonical transformations have also been used to develop effective theories \cite{doi:10.1063/1.4802766, PhysRevA.81.013402, doi:10.1063/1.2196410}.
While useful, it is typically impossible to know whether models built from these techniques accurately describe reality, as measures of model quality are rarely reported.
Instead, out-of-method comparisons of model predictions to experiment are conducted for model validation.

The recently developed DMD \cite{Zheng2017} method provides a more direct link between \textit{ab-initio} and many-body effective Hamiltonians.
Similar to the methods above, DMD uses information from \textit{ab-initio} simulations on the system to fit the best parameters for a given many-body model \textit{ansatz}.
Unlike the methods above, the resultant fit model is accompanied by a clear measure of model quality, quantifying the ability of the model to reproduce the low-energy spectrum of the system.
Since model quality is reported, DMD allows for direct comparison between different model \textit{ansatzes} as well, an ability which can be used to learn the most important interaction terms for a given system directly from \textit{ab-initio} calculations.
DMD has been used to develop accurate many-body Hamiltonians for benzene, a 1-d hydrogen chain, and a preliminary on-site Hubbard model for graphene \cite{Zheng2017, Wagner2015}.

The utility of DMD is made apparent when studying the recent surge in model development for TBLG.
\textit{Ab-initio} model development has so far been restricted to tight-binding models which accurately describe single particle effects like band flattening, but without a report of model quality cannot determine the role of interactions in the system \cite{Fang, Fang2016, PhysRevX.8.031088, PhysRevX.8.031087}.
Minimal interacting models have been proposed using traditional theoretical methods, but without a link to \textit{ab-initio} it is unclear whether these simplified models can capture the complexity of the real system \cite{PhysRevB.98.045103, PhysRevB.100.125116, PhysRevX.8.031089}.
Starting with the tight-binding \textit{ansatz}, DMD can be used to learn the most important interactions in the system, resulting in a final many-body model which can accurately describe TBLG.

As such, one portion of a project between research groups at UIUC, UCSB and Rice University is dedicated towards the development of an accurate continuum-scale interacting model for TBLG.
The model development consists of two layers of downfolding: first, developing an atomic scale lattice model for TBLG using \textit{ab-initio} simulations, and second, downfolding the resultant lattice model to a continuum length scale.
My thesis will be focused on the first layer of downfolding, which I will conduct within the DMD framework using accurate \textit{ab-initio} fixed-node diffusion Monte Carlo calculations.
I propose a two year plan which will result in atomic scale, many-body lattice Hamiltonians for single- (SLG) and simple bi-layer (BLG) graphene systems.
My thesis will lay the necessary ground work for the development of a fully twist angle dependent lattice model for TBLG, and the eventual continuum model described above.
